Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-04-09 phyloglmm_ms.tex
finally reach the discussion .... again...

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
  \caption{Comparsion of single group model parameter estimates across different R packages in Table \ref{table:platform}. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.} 
\label{ssplot}
\end{figure}
\end{center}

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
  \caption{Comparsion of multi group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
  \caption{Comparsion of multi group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot_time}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
  \caption{Comparsion of multi group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot_coverage}
\end{figure}
\end{center}

Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$).
The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters.  
% \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
% \mli{save it for random slope discussion}

Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr}. 
\pkg{glmmTMB} takes $\approx$ 125 seconds for lme4 to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

>>>>

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.} 
\label{ssplot}
\end{figure}
\end{center}

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
  \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
  \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot_time}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
  \caption{Comparison of multi-group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
  \label{msplot_coverage}
\end{figure}
\end{center}

Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$) for few number of species (i.e. n = 25 and 50).
The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters.

Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr}. 
\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
